Compositional Certified Resource Bounds

### 2014-07-01 paper/main.tex
some fixes in the rules
add experiments table
this and that

  The practicability of the analysis framework is demonstrated with an
  experimental evaluation.  The expressivity of the logic is
  demonstrated by manually deriving customized bounds that are
  tailored to specific algorithms.  A comparison of the automatic
  amortized analysis with other automatic tools on 30 challenging
  examples from the literature and open-source software shows that the
  derived bounds are often more precise.
\end{abstract}

@@

To find derivations in the quantitative Hoare logic automatically, we
have to focus on bounds that have a certain shape.  The general form
of \emph{potential functions} (or assertions) that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is again a simplified program state as introduced in \pref{sec:sem},
$|\inter a b| = max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as
$$
  \Phi = \sum_{i \in I} q_i f_i.
$$
We often write $xy$ to denote the index $(x,y)$.  The family
$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense) for
potential functions.  This allows us to uniquely represent any linear
potential function $\Phi$ as a \emph{quantitative annotation} $Q =
(q_i)_{i \in I}$, that is, a family of rational numbers where only a
finite number of elements are not zero.
%
For soundness, we require that potential functions always give
non-negative results.  This is acheived by restricting coefficients
in a quantitative annotation to be always non-negative.

@@

@@

The deductive system defined in~\pref{fig:auto} derives
judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$
These judgements are analogous to the logic judgements
$\Delta; B; R \vdash \htriple P S Q$ where each assertion
splits in two parts, a logical part $\Gamma$ and a quantitative
part $Q$.  We left the function context $\Delta$ appart in
the presentation to avoid cluttering notations, but it is
implicit on all judgements.  We call precondition the pair
$(\Gamma, Q)$ and postcondition the pair $(\Gamma', Q')$.
The correspondence between judgements of the automation
and the ones of the quantitative logic is made formal by our
soundness proof of the analysis.

Since the different rules of~\pref{fig:auto} manage the
logical state in exactly the same way as the quantitative
logic, we will focus on the quantitative annotations in the
sequel.
%
As a convention, if $Q$ is a quantitative annotation we
will write $q_i$ for its $i$th coordinate.  The notation
$Q \pm n$ used in many rules defines a new context $Q'$
such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\,
q'_i = q_i$.  Finally, if a rule mentions $Q$ and $Q'$
and leaves the latter undefined at some index $i$ we
will implicitely assume that $q'_i = q_i$.

%
% Next paragraph is shit.
%

The {\sc Q:Skip} rule simply reuses its postcondition as
precondition.  This is a sound choice since evaluating
a $\code{skip}$ statement never costs any resource and
does not change the program state.  A similar rule is
the {\sc Q:Tick} rule which enforces that the quantitative
annotation of the precondition is at least $M_t(n)$ to
pay for the resource consumption.  Since the program
state is also unchanged in this case, the logical part
$\Gamma$ of the precondition can soundly be reused in
the postcondition.  A slightly different rule is
{\sc Q:Break}, in that case the regular control flow is
broken, so anything that follows \code{break} in the
program will not be executed.  Thus, the postcondition
$(\Gamma', Q')$ can be arbitrary, and the precondition
is taken from the break part of the judgement similarly
to what is done in the logic.

From these three examples alone, one can see that our
automation follows very closely the rules of the logic.
The case of the update statement is more interesting
and specific to linear potential functions.
In the {\sc Q:IncP} rule the program updates a variable
$x$ to add it $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the
change of the program state.  We write $x'$ the value
of $x$ after the assignment ($x + y$). Since $x$ is
the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.
Note that for any $u$, $\inter x u$ will get smaller
with the update,  and if $x' \in \inter x u$ 
we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.
But $|\inter x {x'}| = |\inter 0 y|$, this means that
the potential $q'_{0y}$ in the postcondition can
be increased by $q_{xu}$ under the guard that
$x' \in \inter x u$.  Dually, the interval
$\inter v x$ can get bigger with the update.
And we know that $|\inter v {x'}| \le y + |\inter v x|$,
so we decrease the potential of $\inter 0 y$
by $q_{vx}$ to pay for this change.  Remark that
the rule forces it only for $v \not\in \mathcal U$,
because if not, we know that $x \le v$ so
$|\inter v x| = 0$.

@@

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
%
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
%
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
%
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
%
\and \Rule{Q:Tick}
{}
{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
%
\\ \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\and \Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \Qplusz
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule{Q:IncP}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:IncM}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
%
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\\\ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \Qplusz
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
}
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\and \Rule{Extend}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
%
\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:auto}
\end{figure*}

@@

>>>>

  An experimental evaluation demonstrates the practicability of the
  analysis framework.  The expressivity of the logic is shown by
  manually deriving customized bounds that are tailored to specific
  algorithms.  A comparison of the automatic amortized analysis with
  other automatic tools on 30 challenging examples from the literature
  and open-source software shows that the bounds derived by the
  automatic amortized analysis are often more precise.
\end{abstract}

@@

To find derivations in the quantitative Hoare logic automatically, we
have to focus on bounds that have a certain shape.  The general form
of \emph{potential functions} (or assertions) that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is again a simplified program state as introduced in \pref{sec:sem},
$|\inter a b| = max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as
$$
  \Phi = \sum_{i \in I} q_i f_i.
$$
We often write $xy$ to denote the index $(x,y)$.  The family
$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense) for
potential functions.  This allows us to uniquely represent any linear
potential function $\Phi$ as a \emph{quantitative annotation} $Q =
(q_i)_{i \in I}$, that is, a family of rational numbers where only a
finite number of elements are not zero.
%
For soundness, we require that potential functions always give
non-negative results.  This is acheived by restricting coefficients
in a quantitative annotation to be always non-negative.

@@

\paragraph{Judgements of the Automatic Analysis.}

The inference system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The inference rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$
These judgements correspond to the logic judgements $\Delta; B; R
\vdash \htriple P S Q$ where each assertion splits in two parts, a
logical part $\Gamma$ and a quantitative part $Q$.  That means that
$(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
$(\Gamma_R, Q_R)$ is the postcondition for return statements, and
$(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
simple Hoare triple.  We leave out the function context $\Delta$ in
the presentation to avoid cluttering notations.  Instead we assume a
fixed function context $\Delta$ that is implicit on all judgements.
Function contexts are treated exactly as in the logic.  We call
precondition the pair $(\Gamma, Q)$ and postcondition the pair
$(\Gamma', Q')$.  The correspondence between judgements of the
automation and the ones of the quantitative logic is made formal by
our soundness proof of the automatic amortized analysis at the end of
this section.

As a convention, if $Q$ and $Q'$ are quantitative annotations then we
assume that $Q = (q_i)_{i\in I}$ is a family with elements $q_i$, $Q'
= (q'_i)_{i \in I}$, etc.  The notation $Q \pm n$ used in many rules
defines a new context $Q'$ such that $q'_0 = q_0 \pm n$ and $\forall i
\neq 0 .\, q'_i = q_i$.  We have the implicit side condition that all
rational coefficients are non-negative.  Finally, if a rule mentions
$Q$ and $Q'$ and leaves the latter undefined at some index $i$ we will
implicitly assume that $q'_i = q_i$.  As mentioned, we leave the
logical contexts $\Gamma$ abstract.

We describe the automatic amortized analysis for a subset of
expressions of C light.  Assignments must have the form \code{x \gets
  y} or \code{x \gets x \pm y}.  In the implementation, a Clight
program is converted into this form prior to analysis without changing
the resource cost.  This achieved by using a series of \emph{cost-free
  assignments} that we do not result in additional cost in the
analysis.  Non-linear operations such as \code{x \gets z*y} are simply
handled by assigning zero potential to coefficients like $q_{xa}$ and
$q_{ax}$ that contain $x$.

@@

Most of the rules correspond exactly to the respective rules of the
quantitative Hoare logic.  For example, the rule {\sc Q:Skip} simply
reuses its postcondition as precondition.  This is sound since
evaluating a $\code{skip}$ statement never costs any resources and
does not change the program state.  Similarly, the rule {\sc Q:Tick}
enforces that the quantitative annotation of the precondition is at
least $M_t(n)$ to pay for the resource consumption.  Since the program
state is also unchanged in this case, the logical part $\Gamma$ of the
precondition can soundly be reused in the postcondition.  In the rule
{\sc Q:Break}, where the linear control flow is broken, the
postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
taken from the break part of the judgement similarly as it is done in
the quantitative logic.  The rule {\sc Q:Return} exhibits a slight
difference to the quantitative logic.  While we have a function $\Z
\to \Assn$ to represent return values in the logic, we assume that a
special variable $\ret$ is part of the index set of $Q_R$.  The
substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
$q_{\ret y} = q'_{x y}$ and $q_{y \ret} = q'_{y x}$ for all $y$ and
$q_i'=q_i$ otherwise.

The most interesting rules are {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
  Q:Inc}.

The case of the update statement is more interesting and specific to
linear potential functions.  In the {\sc Q:IncP} rule the program
updates a variable $x$ to add it $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ the value of $x$ after the assignment
($x + y$). Since $x$ is the only variable changed, only intervals of
the form $\inter u x$ and $\inter x u$ will be resized.  Note that for
any $u$, $\inter x u$ will get smaller with the update, and if $x' \in
\inter x u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'}
u|$.  But $|\inter x {x'}| = |\inter 0 y|$, this means that the
potential $q'_{0y}$ in the postcondition can be increased by $q_{xu}$
under the guard that $x' \in \inter x u$.  Dually, the interval
$\inter v x$ can get bigger with the update.  And we know that
$|\inter v {x'}| \le y + |\inter v x|$, so we decrease the potential
of $\inter 0 y$ by $q_{vx}$ to pay for this change.  Remark that the
rule forces it only for $v \not\in \mathcal U$, because if not, we
know that $x \le v$ so $|\inter v x| = 0$.

@@

All rules except of {\sc Q:Weak} and {\sc Q:Relax} are syntax
directed.

@@

\begin{figure*}[t]
\begin{mathpar}
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
\and \Rule{Q:Tick}
{}
{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
\\ \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
\and \Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \Qplusz
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
\and \Rule{Q:IncP}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
\and \Rule{Q:DecP}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
}
\and \Rule{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\\\ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \Qplusz
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
}
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
}
{ Q'   \succeq_\Gamma Q }
\and \Rule{Q:Extend}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:auto}
\end{figure*}

@@

\begin{figure*}
\tiny

\begin{tabular}{r|p{3cm}c|p{3cm}c|c}

File &
\multicolumn{2}{|c|}{Rank} &
\multicolumn{2}{|c|}{Loopus} &
Amortized Automation
\\
\hline

%--------------------------------------------
\texttt{gcd.c} &

$((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
$O(n)$ &

Fail &
&

$|\inter 0 x| + |\inter 0 y|$
\\

%--------------------------------------------
\texttt{kmp.c} &

$((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
$O(n^2)$ &

$\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
$O(n)$ &

$1 + 2 |\inter 0 n|$
\\

%--------------------------------------------
\texttt{qsort.c} &

Fail &
&

Fail &
&

$1 + 2 |\inter 0 {\code{len}}|$
\\

%--------------------------------------------
\texttt{speed\_pldi09\_fig1.c} &

Fail &
&

$\max(n, 0) + \max(n, 0)$ &
$O(n)$ &

$1 + 2 |\inter 0 n|$
\\

%--------------------------------------------
\texttt{speed\_pldi09\_fig4\_2.c} &

$((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
$O(n)$ &

Fail &
&

$1 + 2 |\inter 0 n|$
\\

%--------------------------------------------
\texttt{speed\_pldi09\_fig4\_4.c} &

$((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
$O(n)$ &

Fail &
&

$|\inter 0 n|$
\\

%--------------------------------------------
\texttt{speed\_pldi09\_fig4\_5.c} &

$((((1+1)+(-1+n))+m)+1)+1$ &
$O(n)$ &

Fail &
&

Fail
\\

%--------------------------------------------
\texttt{speed\_pldi10\_ex1.c} &

$(((1+1)+n)+((-1+((1/2){\cdot}n))+((1/2){\cdot}(n{\cdot}n))))+1$ &
$O(n^2)$ &

$\max(n, 0) + \max(0,  \max(n, 0) {\cdot} \max(0, (n + -1)))$ &
$O(n^2)$ &

Fail
\\

%--------------------------------------------
\texttt{speed\_pldi10\_ex3.c} &

$((((1+1)+(-1+n))+(-1+n))+1)+1$ &
$O(n)$ &

$\max(n, 0) + \max(n, 0)$ &
$O(n)$ &

$|\inter 0 n|$
\\

%--------------------------------------------
\texttt{speed\_pldi10\_ex4.c} &

Fail &
&

Fail &
&

$1 + 2 |\inter 0 n|$
\\

%--------------------------------------------
\texttt{speed\_popl10\_fig2\_1.c} &

$(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
$O(n)$ &

$\max(0, (n - x)) + \max(0, (m - y))$ &
$O(n)$ &

$|\inter x n| + |\inter y m|$
\\

%--------------------------------------------
\texttt{speed\_popl10\_fig2\_2.c} &

$(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
$O(n)$ &

$\max(0, (\max(0, (x + 1 + -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
$O(n)$ &

$|\inter x n| + |\inter z n|$
\\

%--------------------------------------------
\texttt{speed\_popl10\_nested\_multiple.c} &

$(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
$O(n^2)$ &

$\max(0, (m - y)) + \max(0, (n - x))$ &
$O(n)$ &

$|\inter x n| + |\inter y m|$
\\

%--------------------------------------------
\texttt{speed\_popl10\_nested\_single.c} &

$((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
$O(n)$ &

$\max(0, min(n, (n + -1))) + \max(0, min(n, (n + -1)))$ &
$O(n)$ &

$|\inter 0 n|$
\\

%--------------------------------------------
\texttt{speed\_popl10\_sequential\_single.c} &

$(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
$O(n)$ &

$\max(n, 0) + \max(n, 0)$ &
$O(n)$ &

$|\inter 0 n|$
\\

%--------------------------------------------
\texttt{speed\_popl10\_simple\_multiple.c} &

$(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
$O(n)$ &

$\max(n, 0) + \max(m, 0)$ &
$O(n)$ &

$|\inter 0 m| + |\inter 0 n|$
\\

%--------------------------------------------
\texttt{speed\_popl10\_simple\_single\_2.c} &

Fail &
&

$\max(n, 0) + \max(m, 0)$ &
$O(n)$ &

$|\inter 0 n| + |\inter 0 m|$
\\

%--------------------------------------------
\texttt{speed\_popl10\_simple\_single.c} &

$(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
$O(n)$ &

$\max(n, 0)$ &
$O(n)$ &

$|\inter 0 n|$
\\

%--------------------------------------------
\texttt{t07.c} &

$(((1+1)+-1)+x)+1$ &
$O(n)$ &

$\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
$O(n)$ &

$1 + 3 |\inter 0 x| + |\inter 0 y|$
\\

%--------------------------------------------
\texttt{t08.c} &

$(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
$O(n)$ &

$\max(0, (\max(0, (y + -9)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
$O(n)$ &

$1.1 |\inter y z| + 0.1 |\inter 0 y|$
\\

%--------------------------------------------
\texttt{t10.c} &

$(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
$O(n)$ &

$\max(0, (x - y))$ &
$O(n)$ &

$|\inter y x|$
\\

%--------------------------------------------
\texttt{t11.c} &

$(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
$O(n)$ &

$\max(0, (n - x)) + \max(0, (m - y))$ &
$O(n)$ &

$|\inter x n| + |\inter y m|$
\\

%--------------------------------------------
\texttt{t13.c} &

$((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
$O(n^2)$ &

$\max(x, 0) + \max(0, (\max(x, 0) + \max(y, 0)))$ &
$O(n)$ &

$2 |\inter 0 x| + |\inter 0 y|$
\\

%--------------------------------------------
\texttt{t15.c} &

$(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
$O(n)$ &

Fail &
&

$|\inter 0 x|$
\\

%--------------------------------------------
\texttt{t16.c} &

$(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
$O(n)$ &

Fail &
&

$101 |\inter 0 x|$
\\

%--------------------------------------------
\texttt{t19.c} &

$(((1+1)+(151+k))+(-100+i))+1$ &
$O(n)$ &

$\max(0, (i + -100)) + \max(0, (k + i + 51))$ &
$O(n)$ &

$50 + |\inter {-1} i| + |\inter 0 k|$
\\

%--------------------------------------------
\texttt{t20.c} &

$((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
$O(n)$ &

$\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
$O(n)$ &

$|\inter x y| + |\inter y x|$
\\

%--------------------------------------------
\texttt{t27.c} &

Fail &
&

$\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
$O(n)$ &

$0.1|\inter n y| + 11 |\inter n 0|$
\\

%--------------------------------------------
\texttt{t28.c} &

$(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
$O(n)$ &

$\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
$O(n)$ &

$|\inter x 0| + |\inter 0 y| + 1002 |\inter y x|$
\\

%--------------------------------------------
\texttt{t30.c} &

$((1+1)+x)+1$ &
$O(n)$ &

$\max(0, (1 + \max(x, y, 0)))$ &
$O(n)$ &

$|\inter 0 x| + |\inter 0 y|$
\\

%--------------------------------------------
\texttt{t37.c} &

Fail &
&

Fail &
&

$3 + 2 |\inter 0 x| + |\inter 0 y|$
\\

%--------------------------------------------
\texttt{t39.c} &

Fail &
&

Fail &
&

$1.33 + 0.67 |\inter z y|$
\\

%--------------------------------------------
\texttt{t46.c} &

Fail &
&

Fail &
&
